### 2023-06-08 paper/localpsf.tex
paper is back under 26 pages

We present an efficient \emph{matrix-free} point spread function (PSF) method for approximating operators $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ that have locally supported non-negative integral kernels. Here $\Omega \subset \mathbb{R}^\gdim$ is a bounded domain, and $L^2(\Omega)'$ is the space of real-valued continuous linear functionals on $L^2(\Omega)$. Such operators appear, for instance, as Hessians in optimization and inverse problems governed by partial differential equations (PDEs)~\cite{AntilEtAl18,BorziSchulz11}, Schur complements in Schur complement methods for solving partial differential equations (PDEs)~\cite{le1997non,saad1999distributed}, Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps), covariance operators in spatial statistics~\cite{ChenStein21,GeogaEtAl20,LindgrenRueLindstrom11}, and blurring operators in imaging~\cite{DenisEtAl11,NagyOleary98}. Here ``matrix-free'' means that we may apply $\Aop$ and its transpose\footnote{Recall that $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(w) = \left(\Aop^T w\right)(u)$ for all $u,w \in L^2(\Omega)$, where $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$, and similar for operations with $\Aop^T$. }, $\Aop^T$, to arbitrary functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad w \mapsto \Aop^T w,
\end{equation} 
via a black box computational procedure, 
but cannot easily access entries of $\Aop$'s integral kernel. Evaluating the maps in~\eqref{eq:eval_maps} may require solving a sub problem that involves PDEs, or performing other costly computations. By ``non-negative integral kernel,'' we mean that entries of the integral kernel are non-negative numbers; this is not the same as positive semi-definiteness of $\Aop$.

@@

In distributed parameter inverse problems governed by PDEs, one seeks to infer an unknown spatially varying parameter field from limited observations of a state variable that depends on the parameter implicitly through the solution of a PDE. 
%Inverse problems are typically ill-posed, in the sense that many different parameter fields may yield predicted observations that agree with the actual observations to within the noise level. 
Conventionally, the inverse problem is formulated using either a deterministic framework~\cite{AsterBorchersThurber18,BorziSchulz11,Vogel02},
%Guy Chavent, Nonlinear Least Squares for Inverse Problems, Springer, 2009.
or a Bayesian probabilistic framework~\cite{KaipioSomersalo05,Stuart10a,Tarantola05}. In the deterministic framework, one solves an optimization problem to find the parameter that best fits the observations, subject to appropriate regularization~\cite{EnglHankeNeubauer96,Vogel02}. In the probabilistic framework, Bayes' theorem combines the observations with prior information to form a posterior distribution over the space of all possible parameter fields, and computations are performed to extract statistical information about the parameter from this posterior. The Hessian of the objective function in the determinstic optimization problem and the Hessian of the negative log posterior in the Bayesian setting are equal or approximately equal to each other under typical noise, regularization, and prior models, so we refer to both of these Hessians as ``the Hessian.'' The Hessian consists of a data misfit term (the \emph{data misfit Hessian}), which depends on a discrepancy between the observations and the associated model predictions, and a regularization or prior term (the \emph{regularization Hessian}) which does not depend on the observations. For more details on the Hessian, see~\cite{Alger19,GhattasWillcox21,VillaPetraGhattas21}.

Hessian approximations and preconditioners are highly desirable because the Hessian is central to efficient solution of inverse problems in both the deterministic and Bayesian settings. When solving the deterministic optimization problem with Newton-type methods, the Hessian is the coefficient operator for the linear system that must be solved or approximately solved at every Newton iteration. Good Hessian preconditioners reduce the number of iterations required to solve these Newton linear systems with Krylov methods such as the conjugate gradient method~\cite{Saad03}. In the Bayesian setting, the inverse of the Hessian is the covariance of a local Gaussian approximation of the posterior. This Gaussian distribution can be used directly as an approximation of the posterior, or it can be utilized as a proposal for Markov chain Monte-Carlo methods for drawing samples from the posterior. For instance, see~\cite{KimEtAl21,PetraEtAl14} and the references therein.

@@

Examples of inverse problems with high rank data misfit Hessians include large-scale ice sheet inverse problems \cite{HartlandEtAl23,IsaacEtAl15}, advection dominated advection-diffusion inverse problems~\cite{AkcelikBirosDraganescuEtAl05}\cite[Chapter 5]{Flath13}, high frequency wave inverse problems~\cite{BuiEtAl13}, and more generally, all inverse problems in which the observations highly inform the parameter. The eigenvalues of the data misfit Hessian characterize the sensitivity of the predicted observations to perturbations of the parameter in the corresponding eigenvector directions,
%(see Figure~\ref{fig:hessian_eigenstructure})
hence more informative data leads to larger eigenvalues and therefore a larger numerical rank~\cite{AlexanderianGloorGhattas16}\cite[Section 1.4 and Chapter 4]{Alger19}. Roughly speaking, the numerical rank of the data misfit Hessian is the dimension of the subspace of parameter space that is informed by the data. The numerical rank of the regularization preconditioned data misfit Hessian may be reduced by increasing the strength of the regularization, but this throws away useful information. If the regularization is too strong, then components of the parameter that could be learned from the observations
%components of the observations that have a large signal to noise ratio 
would instead be reconstructed based on the regularization~\cite[Section 4]{AlgerEtAl17}\cite[Chapters 1 and 7]{Vogel02}. Hence, low rank approximation methods suffer from a predicament: if the data highly inform the parameter and the regularization is chosen appropriately, then a large number of operator applies are required to form an accurate approximation of the Hessian using low rank approximation methods. 
%The best case scenario from a scientific perspective (highly informative data) is the worst case scenario from a computational perspective (large computational cost)~\cite{AlexanderianGloorGhattas16,Alger19,AlgerEtAl17}. 
High rank Hessian approximation methods are needed.

@@

Let $\Omega \subset \mathbb{R}^\gdim$ be a bounded domain (typically $\gdim=1$, $2$, or $3$). We seek to approximate integral operators $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ of the form
\begin{equation}
\label{eq:kernel_representation}
(\Aop u)(w) := \int_\Omega \int_\Omega w(y) \Aker(y,x) u(x) \mathrm{d}x \mathrm{d}y.
\end{equation}
Here the linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, the scalar $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$.
The integral kernel, $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$, exists in principle but is not easily accessible. In this section we describe how to extend the application of $\Aop$ to distributions, which allows us to define impulse responses (Section~\ref{sec:impulse_response_2}), we then state the conditions on $\Aop$ that the proposed method requires (Section~\ref{sec:conditions_2}), and detail finite element discretization (Section~\ref{sec:finite_element_kernel}).

@@

The application of $\Aop$ may be extended to distributions if $\Aker$ is sufficiently regular. 
%We derive the extension as follows. 
Let $\rho \in L^2(\Omega)'$, and let $\rho^* \in L^2(\Omega)$ denote the Riesz representative of $\rho$ with respect to the $L^2(\Omega)$ inner product. We have
\begin{subequations}
	\label{eq:extension_to_distributions}
	\begin{align}
		\left(\Aop \genericdistribution^*\right)(w) &= \int_\Omega \int_\Omega w(y) \Aker(y,x) \genericdistribution^*(x) \mathrm{d}x ~\mathrm{d}y \\
		&= \int_\Omega w(y) \int_\Omega \Aker(y,x) \genericdistribution^*(x) \mathrm{d}x ~\mathrm{d}y 
		= \int_\Omega w(y) \genericdistribution\left(\Aker(y,\cdot)\right) \mathrm{d}y, \label{eq:action_on_distribution}
	\end{align}
\end{subequations}
where $\Aker(y,\cdot)$ denotes the function $x \mapsto \Aker(y,x)$.
Now let $\mathcal{D}(\Omega) \subset L^2(\Omega)$ be a suitable space of test functions and let $\rho:\mathcal{D}(\Omega) \rightarrow \mathbb{R}$ be a distribution. In this case, the Riesz representative $\rho^*$ may not exist, so the derivation in~\eqref{eq:extension_to_distributions} is not valid. However, if $\Aker$ is sufficiently regular such that the function $y \mapsto \rho\left(\Aker(y,~\cdot~)\right)$ is well-defined for almost all $y\in\Omega$, and if this function is in $L^2(\Omega)$, then the right hand side of~\eqref{eq:action_on_distribution} is well-defined. Hence, we \emph{define} the application of $\Aop$ on the distribution $\rho$ to be the right hand side of~\eqref{eq:action_on_distribution}. We denote this operator application by ``$\Aop \genericdistribution^*$,'' even if $\rho^*$ does not exist.

@@

After discretization, the operator $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ is replaced by an operator $A_h:V_h \rightarrow V_h'$, which becomes an operator
\begin{equation*}
\mathbf{A}:\mathbb{R}^\fedim_\massmatrix \rightarrow \mathbb{R}^\fedim_{\massmatrix^{-1}}
\end{equation*}
under the isomorphism discussed above. The proposed method is agnostic to the computational procedure for approximating $\Aop$ with $\mathbf{A}$. What is important is that we do not have direct access to matrix entries $\mathbf{A}_{ij}$. Rather, we have a computational procedure that allows us to compute matrix-vector products $\mathbf{u}\mapsto \mathbf{A}\mathbf{u}$ and $\mathbf{w}\mapsto \mathbf{A}^T\mathbf{w}$, and computing these matrix-vector products is costly. The proposed method is designed to mitigate this cost by performing as few of these matrix-vector products as possible.
Of course, matrix entries can be computed via matrix-vector products as $\mathbf{A}_{ij} = \left(\mathbf{A}\mathbf{e}_j\right)_i$,
where $\mathbf{e}_j=(0,\dots,0,1,0,\dots,0)^T$ is the length $\fedim$ unit vector with one in the $j$th coordinate and zeros elsewhere. But computing the matrix-vector product $\mathbf{e}_j \mapsto \mathbf{A}\mathbf{e}_j$ is costly, and therefore wasteful if we do not use other matrix entries in the $j$th column of $\mathbf{A}$. Hence, methods for approximating $\mathbf{A}$ are computationally intractable if they require accessing scattered matrix entries from many different rows and columns of $\mathbf{A}$.

@@

\paragraph{Lumped mass matrix} At the continuum level, $\Aker$ is assumed to be non-negative. However, entries of $\Aker_h$, involve inverse mass matrices, which typically contain negative numbers. If there are too many negative numbers, or if the negative numbers are too large, the proposed method will yield less accurate approximations and may potentially fail. We therefore recommend replacing the mass matrix, $\massmatrix$, with a positive diagonal \emph{lumped mass} approximation. 
In the numerical results, we use the lumped mass matrix in which the $i$th diagonal entry of the lumped mass matrix is the sum of all entries in the $i$th row of the mass matrix. Other mass lumping techniques may be used~\cite{Hughes12}.

@@

Let $x$ and $x'$ be points in $\Omega$ that are ``close'' to each other, and consider the following approximations:
\begin{align}
\phi_x(y) &\approx \phi_{x'}(y) \label{eq:translate1}\\
\phi_x(y) &\approx \phi_{x'}(y-x+x') \label{eq:translate2}\\
\phi_x(y) &\approx \phi_{x'}\left(y-\spatialmean(x)+\spatialmean(x')\right) \label{eq:translate3}\\
\phi_x(y) &\approx \phi_{x'}\left(y-\spatialmean(x)+\spatialmean(x')\right) \spatialvol(x) / \spatialvol(x'). \label{eq:translate4}
\end{align}
These are four different ways to approximate an impulse response by a nearby impulse response, with each successive approximation building upon the previous ones. The proposed method uses~\eqref{eq:translate4}, which is the most sophisticated of these approximations. Approximation~\eqref{eq:translate1} says that $\phi_x$ can be approximated by $\phi_{x'}$ when $x$ and $x'$ are close. Operators satisfying~\eqref{eq:translate1} can be well approximated via low rank CUR approximation~\cite{MahoneyDrineas09}. However, the required rank in the low rank approximation can be large, which makes algorithms based on~\eqref{eq:translate1} expensive. Operators that satisfy~\eqref{eq:translate2} are called ``locally translation invariant'' because integral kernel entries $\Aker(y,x)$ for such operators are approximately invariant under translation of $x$ and $y$ by the same displacement, i.e., $x \rightarrow x+h$ and $y \rightarrow y+h$. It is straightforward to show that if equality holds in~\eqref{eq:translate2}, then $\Aop$ is a convolution operator. Locally translation invariant operators act like convolutions locally, and can therefore be well approximated by PC approximations.

Approximation~\eqref{eq:translate3} improves upon~\eqref{eq:translate1} and~\eqref{eq:translate2}, and generalizes both. On one hand, if~\eqref{eq:translate1} holds, then $\spatialmean(x) \approx \spatialmean(x')$, and so~\eqref{eq:translate3} holds. On the other hand, translating a distribution translates its mean, so if~\eqref{eq:translate2} holds, then $\spatialmean(x')-\spatialmean(x) \approx x' - x$, so again~\eqref{eq:translate3} holds. But approximation~\eqref{eq:translate3} can hold in situations where neither~\eqref{eq:translate1} nor~\eqref{eq:translate2} holds. For example, because the expected value commutes with affine transformations,~\eqref{eq:translate3} will hold when $\Aop$ is locally translation invariant with respect to a translated and rotated frame of reference, while~\eqref{eq:translate2} will not hold in this case. Additionally,~\eqref{eq:translate3} generalizes to operators $\Aop:L^2(\Omega_1) \rightarrow L^2(\Omega_2)'$ that map between function spaces on different domains $\Omega_1$ and $\Omega_2$, and even operators that map between domains with different spatial dimensions. In contrast,~\eqref{eq:translate2} does not naturally generalize to operators that map between function spaces on different domains, because the formula $y-x+x'$ requires vectors in $\Omega_2$ and $\Omega_1$ to be added together.  We call~\eqref{eq:translate3} ``local mean displacement invariance,'' and illustrate~\eqref{eq:translate3} in Figure~\ref{fig:mean_displacement_invariance}.

@@

The ability to rapidly evaluate approximate kernel entries $\Aker(y,x)$ allows us to construct an H-matrix approximation, $\boldsymbol{\Aker}_H \approx \mathbf{\Aker}$, using the conventional adaptive cross H-matrix construction method (Section~\ref{sec:h_matrix_construction_short}). In this method, one forms low rank approximations of off-diagonal blocks of the matrix by sampling rows and columns of those blocks. We then use H-matrix methods to convert $\boldsymbol{\Aker}_H$ into an H-matrix approximation $\mathbf{A}_H \approx \mathbf{A}$. 
%Schematically the algorithm proceeds as follows:
%\begin{equation*}
%\Aop \longrightarrow \spatialvol, \spatialmean, \spatialcov \longrightarrow \underbrace{\{x_i\}_{i=1}^\nsamplepts}_{\text{as }S_b\text{'s}} \longrightarrow \underbrace{\{\impulseresponse_{x_i}\}_{i=1}^\nsamplepts}_{\text{as }\eta_b\text{'s}} \longrightarrow \boldsymbol{\Aker}_H \longrightarrow \mathbf{A}_H.
%\end{equation*}

@@

\paragraph{(2) Building the H-matrix} 
Classical H-matrix construction techniques
%, which is described in Appendix~\ref{app:h_matrix}, 
require evaluating $O(\hrank \fedim \log \fedim)$ matrix entries of the approximation, where $\hrank$ is the H-matrix rank, i.e,  the maximum rank among the blocks of the H-matrix \cite{BebendorfRjasanow03}. To evaluate one matrix entry, first one must find the $\numnbr$ nearest sample points to a given point, where $\numnbr$ is the number of impulse responses used in the RBF interpolation. This is done using a precomputed kd-tree of sample points, and requires $O(\numnbr \log \nsamplepts)$ floating point and logical elementary operations. Second, one must find the mesh cells that the points $\{z_i\}_{i=1}^{\numnbr}$ reside in. This is done using an AABB tree of mesh cells, and requires $O(\numnbr \log \fedim)$ elementary operations. Third, one must evaluate finite element basis functions on each of those cells, which requires $O(\numnbr)$ elementary operations. Finally, one must solve a $\numnbr \times \numnbr$ linear system to perform the radial basis function interpolation, which requires $O(\numnbr^3)$ elementary operations. Therefore, building the H-matrix requires
%
%
\begin{equation*}
	O\left(\left(\hrank \fedim \log \fedim\right)\left(\numnbr \log \fedim + \numnbr^3\right)\right) \qquad \text{elementary operations}.
\end{equation*}
%
%
%This follows from the above discussion, and the fact that $\fedim \ge \nsamplepts$.

>>>>

We present an efficient \emph{matrix-free} point spread function (PSF) method for approximating operators $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ that have locally supported non-negative integral kernels. Here $\Omega \subset \mathbb{R}^\gdim$ is a bounded domain, and $L^2(\Omega)'$ is the space of real-valued continuous linear functionals on $L^2(\Omega)$. Such operators appear, for instance, as Hessians in optimization and inverse problems governed by partial differential equations (PDEs)~\cite{AntilEtAl18,BorziSchulz11}, Schur complements in Schur complement methods for solving partial differential equations (PDEs)~\cite{le1997non,saad1999distributed}, Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps), covariance operators in spatial statistics~\cite{ChenStein21,GeogaEtAl20,LindgrenRueLindstrom11}, and blurring operators in imaging~\cite{DenisEtAl11,NagyOleary98}. Here ``matrix-free'' means that we may apply $\Aop$ and its transpose\footnote{Recall that $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(w) = \left(\Aop^T w\right)(u)$ for all $u,w \in L^2(\Omega)$, where $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$, and similar for operations with $\Aop^T$. }, $\Aop^T$, to functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad w \mapsto \Aop^T w,
\end{equation} 
via a black box computational procedure, 
but cannot easily access entries of $\Aop$'s integral kernel. Evaluating the maps in~\eqref{eq:eval_maps} may require solving a sub problem that involves PDEs, or performing other costly computations. By ``non-negative integral kernel,'' we mean that entries of the integral kernel are non-negative numbers; this is not the same as positive semi-definiteness of $\Aop$.

@@

In distributed parameter inverse problems governed by PDEs, one seeks to infer an unknown spatially varying parameter field from limited observations of a state variable that depends on the parameter implicitly through the solution of a PDE. 
%Inverse problems are typically ill-posed, in the sense that many different parameter fields may yield predicted observations that agree with the actual observations to within the noise level. 
Conventionally, the inverse problem is formulated using either a deterministic framework~\cite{AsterBorchersThurber18,BorziSchulz11,Vogel02},
%Guy Chavent, Nonlinear Least Squares for Inverse Problems, Springer, 2009.
or a Bayesian probabilistic framework~\cite{KaipioSomersalo05,Stuart10a,Tarantola05}. In the deterministic framework, one solves an optimization problem to find the parameter that best fits the observations, subject to appropriate regularization~\cite{EnglHankeNeubauer96,Vogel02}. In the probabilistic framework, Bayes' theorem combines the observations with prior information to form a posterior distribution over the space of all possible parameter fields, and computations are performed to extract statistical information about the parameter from this posterior. The Hessian of the objective function in the determinstic optimization problem and the Hessian of the negative log posterior in the Bayesian setting are equal or approximately equal under typical noise, regularization, and prior models, so we refer to both of these Hessians as ``the Hessian.'' The Hessian consists of a data misfit term (the \emph{data misfit Hessian}), which depends on a discrepancy between the observations and the associated model predictions, and a regularization or prior term (the \emph{regularization Hessian}) which does not depend on the observations. For more details on the Hessian, see~\cite{Alger19,GhattasWillcox21,VillaPetraGhattas21}.

Hessian approximations and preconditioners are highly desirable because the Hessian is central to efficient solution of inverse problems in both the deterministic and Bayesian settings. When solving the deterministic optimization problem with Newton-type methods, the Hessian is the coefficient operator for the linear system that must be solved or approximately solved at every Newton iteration. Good Hessian preconditioners reduce the number of iterations required to solve these Newton linear systems with the conjugate gradient method~\cite{Saad03}. In the Bayesian setting, the inverse of the Hessian is the covariance of a local Gaussian approximation of the posterior. This Gaussian distribution can be used directly as an approximation of the posterior, or it can be used as a proposal for Markov chain Monte-Carlo methods for drawing samples from the posterior. For instance, see~\cite{KimEtAl21,PetraEtAl14} and the references therein.

@@

Examples of inverse problems with high rank data misfit Hessians include large-scale ice sheet inverse problems \cite{HartlandEtAl23,IsaacEtAl15}, advection dominated advection-diffusion inverse problems~\cite{AkcelikBirosDraganescuEtAl05}\cite[Chapter 5]{Flath13}, high frequency wave inverse problems~\cite{BuiEtAl13}, and more generally, all inverse problems in which the observations highly inform the parameter. The eigenvalues of the data misfit Hessian characterize the sensitivity of the predicted observations to perturbations of the parameter in the corresponding eigenvector directions,
%(see Figure~\ref{fig:hessian_eigenstructure})
hence more informative data leads to larger eigenvalues and a larger numerical rank~\cite{AlexanderianGloorGhattas16}\cite[Section 1.4 and Chapter 4]{Alger19}. Roughly speaking, the numerical rank of the data misfit Hessian is the dimension of the subspace of parameter space that is informed by the data. The numerical rank of the regularization preconditioned data misfit Hessian may be reduced by increasing the strength of the regularization, but this throws away useful information: components of the parameter that could be learned from the observations
%components of the observations that have a large signal to noise ratio 
would instead be reconstructed based on the regularization~\cite[Section 4]{AlgerEtAl17}\cite[Chapters 1 and 7]{Vogel02}. Hence, low rank approximation methods suffer from a predicament: if the data highly inform the parameter and the regularization is chosen appropriately, then a large number of operator applies are required to form an accurate approximation of the Hessian using low rank approximation methods. 
%The best case scenario from a scientific perspective (highly informative data) is the worst case scenario from a computational perspective (large computational cost)~\cite{AlexanderianGloorGhattas16,Alger19,AlgerEtAl17}. 
High rank Hessian approximation methods are needed.

@@

Let $\Omega \subset \mathbb{R}^\gdim$ be a bounded domain (typically $\gdim=1$, $2$, or $3$). We seek to approximate integral operators $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ of the form
\begin{equation}
\label{eq:kernel_representation}
(\Aop u)(w) := \int_\Omega \int_\Omega w(y) \Aker(y,x) u(x) \mathrm{d}x \mathrm{d}y.
\end{equation}
The linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, and the scalar $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$.
The integral kernel, $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$, exists but is not easily accessible. In this section we describe how to extend the domain of $\Aop$ to distributions, which allows us to define impulse responses (Section~\ref{sec:impulse_response_2}), we then state the conditions on $\Aop$ that the proposed method requires (Section~\ref{sec:conditions_2}), and detail finite element discretization (Section~\ref{sec:finite_element_kernel}).

@@

The operator $\Aop$ may be applied to distributions if $\Aker$ is sufficiently regular. 
%We derive the extension as follows. 
Given $\rho \in L^2(\Omega)'$, let $\rho^* \in L^2(\Omega)$ denote the Riesz representative of $\rho$ with respect to the $L^2(\Omega)$ inner product. We have
\begin{subequations}
	\label{eq:extension_to_distributions}
	\begin{align}
		\left(\Aop \genericdistribution^*\right)(w) &= \int_\Omega \int_\Omega w(y) \Aker(y,x) \genericdistribution^*(x) \mathrm{d}x ~\mathrm{d}y \\
		&= \int_\Omega w(y) \int_\Omega \Aker(y,x) \genericdistribution^*(x) \mathrm{d}x ~\mathrm{d}y 
		= \int_\Omega w(y) \genericdistribution\left(\Aker(y,\cdot)\right) \mathrm{d}y, \label{eq:action_on_distribution}
	\end{align}
\end{subequations}
where $\Aker(y,\cdot)$ denotes the function $x \mapsto \Aker(y,x)$.
Now let $\mathcal{D}(\Omega) \subset L^2(\Omega)$ be a suitable space of test functions and let $\rho:\mathcal{D}(\Omega) \rightarrow \mathbb{R}$ be a distribution. In this case, 
%the Riesz representative 
$\rho^*$ may not exist, so the derivation in~\eqref{eq:extension_to_distributions} is not valid. However, if $\Aker$ is sufficiently regular such that the function $y \mapsto \rho\left(\Aker(y,~\cdot~)\right)$ is well-defined for almost all $y\in\Omega$, and if this function is in $L^2(\Omega)$, then the right hand side of~\eqref{eq:action_on_distribution} is well-defined. Hence, we \emph{define} the application of $\Aop$ to the distribution $\rho$ to be the right hand side of~\eqref{eq:action_on_distribution}. We denote this operator application by ``$\Aop \genericdistribution^*$,'' even if $\rho^*$ does not exist.

@@

After discretization, the operator $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ is replaced by an operator $A_h:V_h \rightarrow V_h'$, which becomes an operator
$
\mathbf{A}:\mathbb{R}^\fedim_\massmatrix \rightarrow \mathbb{R}^\fedim_{\massmatrix^{-1}}
$
under the isomorphism discussed above. The proposed method is agnostic to the computational procedure for approximating $\Aop$ with $\mathbf{A}$. What is important is that we do not have direct access to matrix entries $\mathbf{A}_{ij}$. Rather, we have a computational procedure that allows us to compute matrix-vector products $\mathbf{u}\mapsto \mathbf{A}\mathbf{u}$ and $\mathbf{w}\mapsto \mathbf{A}^T\mathbf{w}$, and computing these matrix-vector products is costly. The proposed method mitigates this cost by performing as few matrix-vector products as possible.
Of course, matrix entries can be computed via matrix-vector products as $\mathbf{A}_{ij} = \left(\mathbf{A}\mathbf{e}_j\right)_i$,
where $\mathbf{e}_j=(0,\dots,0,1,0,\dots,0)^T$ is the length $\fedim$ unit vector with one in the $j$th coordinate and zeros elsewhere. But computing the matrix-vector product $\mathbf{e}_j \mapsto \mathbf{A}\mathbf{e}_j$ is costly, and therefore wasteful if we do not use other matrix entries in the $j$th column of $\mathbf{A}$. Hence, methods for approximating $\mathbf{A}$ are computationally intractable if they require accessing scattered matrix entries from many different rows and columns of $\mathbf{A}$.

@@

\paragraph{Lumped mass matrix} At the continuum level, $\Aker$ is assumed to be non-negative. However, entries of $\Aker_h$, involve inverse mass matrices, which typically contain negative numbers. If there are too many negative numbers, or if the negative numbers are too large, the proposed method will yield less accurate approximations and may potentially fail. We therefore recommend replacing the mass matrix, $\massmatrix$, with a positive diagonal \emph{lumped mass} approximation. 
Here, we use the lumped mass matrix in which the $i$th diagonal entry of the lumped mass matrix is the sum of all entries in the $i$th row of the mass matrix. Other mass lumping techniques may be used~\cite{Hughes12}.

@@

Let $x$ and $x'$ be points in $\Omega$ that are ``close'' to each other, and consider the following approximations:
\begin{align}
\phi_x(y) &\approx \phi_{x'}(y) \label{eq:translate1}\\
\phi_x(y) &\approx \phi_{x'}(y-x+x') \label{eq:translate2}\\
\phi_x(y) &\approx \phi_{x'}\left(y-\spatialmean(x)+\spatialmean(x')\right) \label{eq:translate3}\\
\phi_x(y) &\approx \phi_{x'}\left(y-\spatialmean(x)+\spatialmean(x')\right) \spatialvol(x) / \spatialvol(x'). \label{eq:translate4}
\end{align}
These are four different ways to approximate an impulse response by a nearby impulse response, with each successive approximation building upon the previous ones. The proposed method uses~\eqref{eq:translate4}, which is the most sophisticated. Approximation~\eqref{eq:translate1} says that $\phi_x$ can be approximated by $\phi_{x'}$ when $x$ and $x'$ are close. Operators satisfying~\eqref{eq:translate1} can be well approximated via low rank CUR approximation~\cite{MahoneyDrineas09}. However, the required rank in the low rank approximation can be large, which makes algorithms based on~\eqref{eq:translate1} expensive. Operators that satisfy~\eqref{eq:translate2} are called ``locally translation invariant'' because integral kernel entries $\Aker(y,x)$ for such operators are approximately invariant under translation of $x$ and $y$ by the same displacement, i.e., $x \rightarrow x+h$ and $y \rightarrow y+h$. It is straightforward to show that if equality holds in~\eqref{eq:translate2}, then $\Aop$ is a convolution operator. Locally translation invariant operators act like convolutions locally, and can therefore be well approximated by PC approximations.

Approximation~\eqref{eq:translate3} improves upon~\eqref{eq:translate1} and~\eqref{eq:translate2}, and generalizes both. On one hand, if~\eqref{eq:translate1} holds, then $\spatialmean(x) \approx \spatialmean(x')$, and so~\eqref{eq:translate3} holds. On the other hand, translating a distribution translates its mean, so if~\eqref{eq:translate2} holds, then $\spatialmean(x')-\spatialmean(x) \approx x' - x$, so again~\eqref{eq:translate3} holds. But approximation~\eqref{eq:translate3} can hold in situations where neither~\eqref{eq:translate1} nor~\eqref{eq:translate2} holds. For example, because the expected value commutes with affine transformations,~\eqref{eq:translate3} will hold when $\Aop$ is locally translation invariant with respect to a translated and rotated frame of reference, while~\eqref{eq:translate2} will not. Additionally,~\eqref{eq:translate3} generalizes to operators $\Aop:L^2(\Omega_1) \rightarrow L^2(\Omega_2)'$ that map between function spaces on different domains $\Omega_1$ and $\Omega_2$, and even operators that map between domains with different spatial dimensions. In contrast,~\eqref{eq:translate2} does not naturally generalize to operators that map between function spaces on different domains, because the formula $y-x+x'$ requires vectors in $\Omega_2$ and $\Omega_1$ to be added together.  We call~\eqref{eq:translate3} ``local mean displacement invariance,'' and illustrate~\eqref{eq:translate3} in Figure~\ref{fig:mean_displacement_invariance}.

@@

The ability to rapidly evaluate approximate kernel entries $\Aker(y,x)$ allows us to construct an H-matrix approximation, $\boldsymbol{\Aker}_H \approx \mathbf{\Aker}$, using the conventional adaptive cross H-matrix construction method (Section~\ref{sec:h_matrix_construction_short}). In this method, one forms low rank approximations of off-diagonal blocks of the matrix by sampling rows and columns of those blocks. We then convert $\boldsymbol{\Aker}_H$ into an H-matrix approximation $\mathbf{A}_H \approx \mathbf{A}$. 
%Schematically the algorithm proceeds as follows:
%\begin{equation*}
%\Aop \longrightarrow \spatialvol, \spatialmean, \spatialcov \longrightarrow \underbrace{\{x_i\}_{i=1}^\nsamplepts}_{\text{as }S_b\text{'s}} \longrightarrow \underbrace{\{\impulseresponse_{x_i}\}_{i=1}^\nsamplepts}_{\text{as }\eta_b\text{'s}} \longrightarrow \boldsymbol{\Aker}_H \longrightarrow \mathbf{A}_H.
%\end{equation*}

@@

\paragraph{(2) Building the H-matrix} 
Classical H-matrix construction techniques
%, which is described in Appendix~\ref{app:h_matrix}, 
require evaluating $O(\hrank \fedim \log \fedim)$ matrix entries of the approximation \cite{BebendorfRjasanow03}, where $\hrank$ is the H-matrix rank, i.e,  the maximum rank among the blocks of the H-matrix. To evaluate one matrix entry, first one must find the $\numnbr$ nearest sample points to a given point, where $\numnbr$ is the number of impulse responses used in the RBF interpolation. This is done using a precomputed kd-tree of sample points, and requires $O(\numnbr \log \nsamplepts)$ floating point and logical elementary operations. Second, one must find the mesh cells that the points $\{z_i\}_{i=1}^{\numnbr}$ reside in. This is done using an AABB tree of mesh cells, and requires $O(\numnbr \log \fedim)$ elementary operations. Third, one must evaluate finite element basis functions on those cells, which requires $O(\numnbr)$ elementary operations. Finally, the radial basis function interpolation requires solving a $\numnbr \times \numnbr$ linear system, which requires $O(\numnbr^3)$ elementary operations. Therefore, building the H-matrix requires
%
%
\begin{equation*}
	O\left(\left(\hrank \fedim \log \fedim\right)\left(\numnbr \log \fedim + \numnbr^3\right)\right) \qquad \text{elementary operations}.
\end{equation*}
%
%
%This follows from the above discussion, and the fact that $\fedim \ge \nsamplepts$.
